---
title: "Rbootcamp"
author: "Michael Kleinsasser"
date: "8/8/2019"
output: 
    beamer_presentation:
      theme: "CambridgeUS"
      colortheme: "dolphin"
      fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Personal Introduction

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}

\begin{itemize}
\setlength\itemsep{1em}

\item R programmer for the Department of Biostatistics

\item Write and mantain R packages for faculty and students

\item Consult faculty and students on writing R packages

\end{itemize}


\end{column}%
\hfill%
\begin{column}{.48\textwidth}

\includegraphics[width = 2in]{personal_picture.jpg}

\end{column}%
\end{columns}

## Rbootcamp Introduction

Goals:

- How to learn R

- Data manipulation with dplyr

- Reproducible research

- Simulation

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

